PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-02-13 papers/NAR_web_server_2009/paper.tex
Added workflow figure

\history{Received Feb. 12, 2010}

@@

\begin{abstract}
  We present {\sc Peace}, a standalone tool for the high-throughput
  {\it ab initio} clustering of transcript fragment sequences produced
  by Next Generation or Sanger Sequencing technologies.  Installed and
  run through an easily downloaded GUI, {\sc Peace} can process large
  data sets and group the fragments by gene association, achieving a
  greater sensitivity (without sacrificing runtime) than the competing
  clustering tools.  Through the GUI the user can collect
  cluster statistics and examine specific clusters for more
  comprehensive study.  {\sc Peace}, used as a ``pre-assembly'' tool,
  enables any assembly tool to work with individual small clusters of
  closely related sequences, resulting in signficant increases in
  sensitivit and robustness to sequenceing error, as well as a reduction in
  overall runtime.
\end{abstract}

@@

Understanding an organism's transcriptome, the set of (spliced)
transcripts expressed by genes of the organism, is a vital step in
understanding the full functional and organizational role of the
genome in the life cycle of any eukaryote.  Studying the transcriptome
has led to gene discovery, provided information on splice variants,  
and helped shed light on the biological processes both controlling and
dictated by the genome \cite{Nagaraj07}.  However, to access to those
transcripts, we must deal with the fragmented data produced by
both Next Generation and traditional Sanger sequencing technology.

Until recently, access to a transcriptome sequence was primarily
through the use of Expressed Sequence Tags (ESTs), single-pass cDNA
sequences derived from transcribed mRNAs and sequenced by Sanger
Sequencing technology.  More recently, Next Generation Sequencing
(NGS) technology has begun to rapidly replace Sanger Sequencing,
allowing for more comprehensive coverage of the transcriptome.  For
example, ESTs now being added to the GenBank dbEST are increasingly
the product of NGS technologies such as 454 pyrosequencing, which
enables the sequencing of novel and rare transcripts at a considerably
higher rate of coverage than Sanger Sequencing
\cite{Cheung2006,Emrich2007}.  From a computational perspective, this
is a mixed blessing: while NGS provides immense quantities of new
information, it also provides immensely larger data sets, requiring
faster computational techniques if we are to process them.

Given a set of transcript fragments sampled from across the genome, a
necessary first step in the analysis is that of clustering: separating
the fragments according to the transcript from which they were
derived.  Frequently performed implicitly by assembly tools,
clustering the data as a ``pre-assembly'' step has a number of
advantages.  Most significantly: performing this step will
allow the application of the assembly tool to individual clusters --
saving significant amounts of time due to the reduced input size
\cite{Hazelhurst08a}.  But clustering is a computationally challenging
problem.  Even with the smaller number of ESTs produced using Sanger
Sequencing, the runtime and memory requirements to cluster based on
pair-wise sequence alignments make such an approach infeasible in practice.  The
much larger data-set size produced by NGS 
technologies exacerbates this problem.

Here we present {\sc Peace}, a tool for the {\it ab initio} clustering
of transcript fragments by gene, applicable to both NGS and
traditional Sanger Sequencing technologies.  Available through the
\href{http://www.peace-tools.org}{www.peace-tools.org} website, the
{\sc Peace} GUI allows the user to both easily install (locally or remotely)
and run the computational tool, as well as enabling transparent
parallel processing and providing various tools for result analysis.
Using a {\it minimum spanning tree} (MST) based clustering strategy
\cite{Jain99,Wan08} and the $d^2$ sequence distance function
\cite{Hide94}, {\sc Peace} generates the MSTs using a tailored version
of Prim's algorithm \cite{Prim57}.  The result is an easy-to-use and
clustering tool that, without any sacrifice in runtime, is
considerably more sensitive and robust to sequencing errors than other
clusters tools in the literature \cite{Burke99,Slater00,
  Huang99,Parkinson02,Kalyanaraman03,Malde03,Ptitsyn05,Hazelhurst08a, Picardi09}.

@@

The {\sc Peace} GUI can be launched by downloading and executing the
Jar file available on the {\sc Peace} wesite
(\href{http://www.peace-tools.org}{www.peace-tools.org}) to any
machine running a standard Java Virtual Machine (JVM).  Once running,
the user can employ the GUI to install the computational tool and perform
a clustering of a data file in the FASTA format, view an initial
analysis of the clusters, and produce files containing subsets of the
clusters for assembly by a tool such as {\sc Cap3} \cite{Huang99}.
A typical (first) use of {\sc Peace} generally
requires the the following steps.

@@

\noindent {\bf Result Analysis:} Once the resulting clusters have been
computed, the user has several options for analysis:
\begin{itemize}
\item {\bf Export:} The user can export the contents of one or more
  clusters into a FASTA format file, obtaining a subset of the original target file
  containing the sequences corresponding to the selected clusters
  ready for processing by an assembly tool (e.g. {\sc Cap3} \cite{Huang99}).
\item {\bf View Clustering}: The user may view a list of clusters,
  expanding selected clusters to a list of all individual sequences (illustrated
  in Figure~1(d)).
\item {\bf Classified Summary Graph:} The user may view a distribution
  of cluster sizes.  Further, the user may set up a {\it classifier},
  associating certain patterns with specific colors.  These patterns
  will be matched against the fragment header information from the
  original FASTA file, allowing the overlay of a colored cluster size
  distributions.  For example, if the sequence names contain unique
  string patterns denoting different cDNA libraries, the classifier
  can help the user to determine and visualize the differential expression profiles
  of different libraries for a given cluster.  The method of setting up these classifiers, and the
  resulting histogram, is illustrated in Figure~1(e).
\end{itemize}
Extensive documentation for the tool has been posted on the {\sc
  Peace} website, as well as links to several tutorial videos demonstrating
{\sc Peace} use and capabilities.

@@

We can model the fragment input as a weighted, undirected graph: the fragments
are represented as nodes, with $d^2$ sequence distances assigned to
the connecting edges as weights.  Conceptually, we want to remove each
edge exceeding a threshold score from the complete graph and define
our partitions by the remaining connected components of the graph.  An
edge with a large weight connects fragments which are likely unrelated;
once such edges are removed the components define a series of
overlaps.  Those fragments that can still be connected by some path
correspond to the same gene.  However, such an approach requires the
calculation of all edge weights.  Such a task is infeasible both in terms of
runtime and memory usage for the data set sizes we expect to process.

We approach the problem by generating a minimum spanning tree of the
described graph, then removing edges exceeding our threshold.  By
using Prim's algorithm we are able to calculate edge weights
on-the-fly (reducing memory requirements) and we can skip the
calculation of a majority of edge distances using the $u/v$ and $t/v$
filtering heuristics employed in {\sc wcd} \cite{Hazelhurst08a}.
These heuristics allow us to quickly dismiss many of the edges as too
large without the need to apply the full $d^2$ algorithm (see
Sections~A and B of the Supplementary Materials for more details).

@@

We have tested {\sc Peace} on both simulated and real data,
comparing results against those produced by the {\sc wcd}
clustering tool \cite{Hazelhurst08a} and the {\sc Cap3} assembly
tool \cite{Huang99} (which calculates a cluster as an implicit, intermediate
step).  For our simulation tests we used the {\bf ESTSim} tool
\cite{Hazelhurst03} to generate simulated transcript of varying length
fragments under different models
of error (Supplementary Materials, Section~C.1),
generating the fragments from the list of 100 zebra fish genes used in
the {\sc wcd} testing \cite{Hazelhurst08a}.  Tool parameters were
taken to match, as closely as possible, those used in the {\sc wcd}
study (see supplementary materials).  Our primary methods of quality
assessment were {\it sensitivity} (the fraction of fragment pairs from the
same gene that were correctly clustered together) and {\it Type 1
  error} (the fraction of genes that were divided between clusters)
\cite{Wang04,Hazelhurst08a}; other measurements are also discussed
in the supplementary materials.  In Figure~2(left) we see a
significant improvement of {\sc Peace} (blue) over {\sc wcd}
(green) and {\sc Cap3} (black) in its sensitivity to sequencing
errors, while in Figure~2(right) we see a comparable
improvement in Type 1 error.  In Figure~3 we look at
the number of singleton clusters (fragments not joined to any cluster),
which will not occur in our simulated sets; we again see significant
improvement in {\sc Peace}.  The sequential runtime of
{\sc Peace} is slightly improved over that of {\sc wcd} (see
Supplementary Materials, Figure~S4).

@@

In applying the tools to real data, we started with a set of
approximately $190,000$ clean Sanger ESTs derived from the {\it
  Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
fragment set to the genome, taking this as our reference clustering.
We see slight improvements in {\sc Peace} over {\sc wcd} in both
sensitivity and Type 1 error (both significantly outperforming {\sc
  Cap3}).  Using the mouse dataset used in Hazelhurst {\it et
  al.}  \cite{Hazelhurst08a} {\sc Peace} again shows slight
improvements in sensitivity and Type 1 error rates, with an 18\% speedup
for {\sc Peace} (see Supplementary Materials,
Section~\href{real_results}, for more details).

@@

Here we have presented {\sc Peace}, a stand-alone tool for the
high-throughput clustering of transcript fragments, managed through a
user friendly GUI that enables both local and remote installation and
execution in sequential or parallel mode.  Based on a novel algorithm
for the clustering of the fragments by gene association, {\sc Peace}
shows significant improvement in sensitivity, without sacrificing runtime,
over the competing {\sc WCD} tool \cite{Hazelhurst08a}, and shows an
order of magnitude in improvement over the clustering preformed in the
course of assembly by the {\sc Cap3} tool \cite{Huang99}.

As a clustering tool based on sequence distance, {\sc Peace} faces
certain inherent limitations. For example, {\sc Peace} cannot handle
duplicate genes; like {\sc WCD}, it is unable to separate clusters
corresponding to genes with a greater than 88\% similarity.
Similarly, natural biological effects such as the trans-splicing of
transcripts, or effects from poorly cleaned transcript data (e.g. the
failure to remove sequencing adapters), post-transcriptional
poly(A)/(T) tails, and low-complexity repeats can cause similar
effects in these clustering tools.  The problems can be handled
through the application of the assembler, and the ability to apply any
assembler to small cluster (as opposed to the data set as a whole)
results in a significant reduction in overall assembly time.

>>>>

\history{Received Feb. 13, 2010}

@@

\begin{abstract}
  We present {\sc Peace}, a stand alone tool for the high-throughput
  {\it ab initio} clustering of transcript fragment sequences produced
  by Next Generation or Sanger Sequencing technologies.  Installed and
  run through an easily downloaded GUI, {\sc Peace} can process large
  data sets and group the fragments by gene association, achieving a
  greater sensitivity (without sacrificing runtime) than the competing
  clustering tools.  Through the GUI the user can collect
  cluster statistics and examine specific clusters for more
  comprehensive study.  {\sc Peace}, used as a ``pre-assembly'' tool,
  enables any assembly tool to work with individual small clusters of
  closely related sequences, resulting in significant increases in
  sensitive and robustness to sequencing error, as well as a reduction in
  overall runtime.
\end{abstract}

@@

Understanding an organism's transcriptome, the set of (spliced)
transcripts expressed by genes of the organism, is a vital step in
understanding the full functional and organizational role of the
genome in the life cycle of any eukaryote.  Studying the transcriptome
has led to gene discovery, provided information on splice variants,  
and helped shed light on the biological processes both controlling and
dictated by the genome \cite{Nagaraj07}.  However, to access those
transcripts, we must deal with the fragmented data produced by
both Next Generation and traditional Sanger sequencing technology.

Until recently, access to a transcriptome sequence was primarily
through the use of Expressed Sequence Tags (ESTs), single-pass cDNA
sequences derived from transcribed mRNAs and sequenced by Sanger
Sequencing technology.  More recently, Next Generation Sequencing
(NGS) technology has begun to rapidly replace Sanger Sequencing,
allowing for more comprehensive coverage of the transcriptome.  For
example, ESTs now being added to the GenBank dbEST are increasingly
the product of NGS technologies such as 454 pyrosequencing, which
enables the sequencing of novel and rare transcripts at a considerably
higher rate of coverage than Sanger Sequencing
\cite{Cheung2006,Emrich2007}.  From a computational perspective, this
is a mixed blessing: while NGS provides immense quantities of new
information, it also provides immensely larger data sets -- and thus a
need for faster, more efficient analysis algorithms.

Given a set of transcript fragments sampled from across the genome, a
necessary first step of their analysis is that of clustering: separating
the fragments according to the transcript from which they were
derived.  Frequently performed implicitly by assembly tools,
clustering the data as a ``pre-assembly'' step has a number of
advantages.  Most significantly: performing this step will
allow the application of the assembly tool to individual clusters --
saving significant amounts of time due to the reduced input size
\cite{Hazelhurst08a}.

But clustering is a computationally challenging problem.  Even with
the smaller number of ESTs produced using Sanger Sequencing, the
runtime and memory requirements to cluster based on pair-wise sequence
alignments make such an approach infeasible in practice.  The much
larger data-set size produced by NGS technologies exacerbates this
problem.  To deal with this, {\sc Peace} combines the $d^2$
alignment-free sequence distance function \cite{Hide94} and the
concept of a {\it minimum spanning tree} \cite{Prim57} to quickly and
accurately find clusters of ESTs expressed from the same gene without
reference to a sequenced genome.  Compared against competing tools
for the same purpose \cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
  Ptitsyn05,Hazelhurst08a,Picardi09}, {\it Peace} proves to be both more sensitive and
more robust to sequencing error without sacrificing runtime.

In short, {\sc Peace} is computational tool for the {\it ab initio}
clustering of transcript fragments by gene association, applicable to both NGS and
traditional Sanger Sequencing technologies.  Available through the
\href{http://www.peace-tools.org}{www.peace-tools.org} website, the
{\sc Peace} GUI allows the user to both easily install (locally or
remotely) and run the clustering engine, as well as enabling
transparent parallel processing and providing various tools for result
analysis.

@@

The {\sc Peace} GUI can be launched by downloading and executing the
Jar file available on the {\sc Peace} website
(\href{http://www.peace-tools.org}{www.peace-tools.org}) to any
machine running the standard Java Virtual Machine (JVM).  Once running,
the user can employ the GUI to install the computational tool and perform
a clustering of a data file in the FASTA format, view an initial
analysis of the clusters, and produce files containing subsets of the
clusters as input to assembly tools such as {\sc Cap3} \cite{Huang99}.
A typical (first) use of {\sc Peace} generally
requires the following steps.

@@

\noindent {\bf Result Analysis:} Once the resulting clusters have been
computed, the user has several options for analysis:
\begin{itemize}
\item {\bf Export:} The user can export the contents of one or more
  clusters into a FASTA format file, obtaining a subset of the original target file
  containing the sequences corresponding to the selected clusters
  ready for processing by an assembly tool (e.g. {\sc Cap3} \cite{Huang99}).
\item {\bf View Clustering}: The user may view a list of clusters,
  expanding selected clusters to a list of all individual sequences (illustrated
  in Figure~1(d)).
\item {\bf Classified Summary Graph:} The user may view a distribution
  of cluster sizes.  Further, the user can set up a {\it classifier},
  associating certain patterns with specific colors.  These patterns
  ware matched against the fragment header information from the
  original FASTA file, allowing the overlay of a colored cluster size
  distributions.  For example, if the sequence names contain unique
  string patterns denoting different cDNA libraries, the classifier
  can help the user to determine and visualize the differential expression profiles
  of different libraries for a given cluster.  The method of setting up these classifiers, and the
  resulting histogram, is illustrated in Figure~1(e).
\end{itemize}
Extensive documentation for the tool has been posted on the {\sc
  Peace} website, as well as links to several tutorial videos demonstrating
{\sc Peace} use and capabilities.

@@

Fragment input is modeled as a weighted, undirected graph: the fragments
are represented as nodes, with $d^2$ sequence distances assigned to
the connecting edges as weights.  Conceptually, we want to remove each
edge exceeding a threshold score from the complete graph and define
our partitions by the remaining connected components of the graph.  An
edge with a large weight connects fragments which are likely unrelated;
once such edges are removed the components define a series of
overlaps.  Those fragments that can still be connected by some path
correspond to the same gene.  However, such an approach requires the
calculation of all edge weights.  Such a task is infeasible both in terms of
runtime and memory usage for the data set sizes we expect to process.

{\sc Peace} approaches the problem by generating a minimum spanning tree of the
described graph, then removing edges exceeding our threshold.  By
using Prim's algorithm we are able to calculate edge weights
on-the-fly (reducing memory requirements) and can skip the
calculation of a majority of edge distances using the $u/v$ and $t/v$
filtering heuristics employed in {\sc wcd} \cite{Hazelhurst08a}.
These heuristics allow us to quickly dismiss many of the edges as too
large without the need to apply the full $d^2$ algorithm (see
Sections~A and B of the Supplementary Materials for more details).

@@

{\sc Peace} has been test on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the {\sc wcd} clustering tool \cite{Hazelhurst08a} and the
{\sc Cap3} assembly tool \cite{Huang99} (which calculates a cluster as
an implicit, intermediate step).  For our simulation tests we used the
{\bf ESTSim} tool \cite{Hazelhurst03} to generate simulated transcript
of varying length fragments under different models of error
(Supplementary Materials, Section~C.1), generating the fragments from
the list of 100 zebra fish genes used in the {\sc wcd} testing
\cite{Hazelhurst08a}.  Tool parameters were taken to match, as closely
as possible, those used in the {\sc wcd} study (see supplementary
materials).  Our primary methods of quality assessment were {\it
  sensitivity} (the fraction of fragment pairs from the same gene that
were correctly clustered together) and {\it Type 1 error} (the
fraction of genes that were divided between clusters)
\cite{Wang04,Hazelhurst08a}; other measurements are also discussed in
the supplementary materials.  In Figure~2(left) we see a significant
improvement of {\sc Peace} (blue) over {\sc wcd} (green) and {\sc
  Cap3} (black) in its sensitivity to sequencing errors, while in
Figure~2(right) we see a comparable improvement in Type 1 error.  In
Figure~3 we look at the number of singleton clusters (fragments not
joined to any cluster), which will not occur in our simulated sets; we
again see significant improvement in {\sc Peace}.  The sequential
runtime of {\sc Peace} is slightly improved over that of {\sc wcd}
(see Supplementary Materials, Figure~S4).

@@

In applying the tools to real data, we started with a set of
approximately $190,000$ clean Sanger ESTs derived from the {\it
  Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
fragment set to the genome, taking this as our reference clustering.
We see slight improvements in {\sc Peace} over {\sc wcd} in both
sensitivity and Type 1 error (both significantly outperforming {\sc
  Cap3}).  Using the mouse data set used in Hazelhurst {\it et al.}
\cite{Hazelhurst08a} {\sc Peace} again shows slight improvements in
sensitivity and Type 1 error rates, with an 18\% speedup for {\sc
  Peace} (see Supplementary Materials, Section~D, for more details).

@@

Here we have presented {\sc Peace}, a stand alone tool for the
high-throughput clustering of transcript fragments, managed through a
user friendly GUI that enables both local and remote installation and
execution in sequential or parallel mode.  Based on a novel algorithm
for the clustering of the fragments by gene association, {\sc Peace}
shows significant improvement in sensitivity, without sacrificing runtime,
over the competing {\sc WCD} tool \cite{Hazelhurst08a}, and shows an
order of magnitude in improvement over the clustering preformed in the
course of assembly by the {\sc Cap3} tool \cite{Huang99}.

As a clustering tool based on sequence distance, {\sc Peace} faces
certain inherent limitations. For example, {\sc Peace} cannot handle
duplicate genes; like {\sc WCD}, it is unable to separate clusters
corresponding to genes with a greater than 88\% similarity.
Similarly, natural biological effects such as the trans-splicing of
transcripts, or effects from poorly cleaned transcript data (e.g. the
failure to remove sequencing adapters or post-transcriptional
poly(A)/(T) tails), and low-complexity repeats can cause similar
effects in these clustering tools.  The problems can be handled
through the application of the assembler, and the ability to apply any
assembler to small cluster (as opposed to the data set as a whole)
results in a significant reduction in overall assembly time.
